### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: 'user' evokes an interactive setting. Go back to the less frightening 'developer' Joe.

Whereas simply-typed languages feature a convenient grammar for
defining data-types and manipulating them, our proposal could be seen
as a step backward. In our setting, the user would have to write a
description instead of a sugared data-type definition. Then, to
recognise an element of $\Nat$ for instance, she would have to
identify $\Mu{}$ of a description as indeed the fix-point of the
pattern functor of natural numbers. Finally, she would have to write
codes instead of the natural data-type constructor.

@@

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to tamper
the second issue. Namely, our user would rather be presented a name
rather than a fix-point of a pattern functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-provided name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

@@

Hence, in the high-level expression language, the user types a
familiar type constructors, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
look-up the tag and turn the constructor form into the appropriate
code.

>>>>

Whereas simply-typed languages feature a convenient grammar for
defining data-types and manipulating them, our proposal could be seen
as a step backward. In our setting, the developer would have to write
a description instead of a sugared data-type definition. Then, to
recognise an element of $\Nat$ for instance, she would have to
identify $\Mu{}$ of a description as indeed the fix-point of the
pattern functor of natural numbers. Finally, she would have to write
codes instead of the natural data-type constructor.

@@

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to tamper
the second issue. Namely, the developer would rather be presented a
name rather than a fix-point of a pattern functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-defined name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

@@

Hence, in the high-level expression language, the developer writes a
familiar type constructors, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
look-up the tag and turn the constructor form into the appropriate
code.

### 2010-03-29 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: 'user' evokes an interactive setting. Go back to the less frightening 'developer' Joe.

The entire work of the magician stands here, in this extension. One
could be worried by the availability of such operations in our type
theory. Indeed, it would be inconvenient for the user to have to
decide between two variants. The solution to this problem is, once
again, to use the type propagation system. Just as the $\F{switch}$
operator, $\F{switchD}$ can be automatically elaborated. To this end,
we extend the type-checking rules (Fig.~\ref{fig:type-checking}) with
the following inference rule:

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}$
completes the implementation. To sum up, we have implemented the free
monad construction for an arbitrary tagged description. This gives our
user the ability, for any data-type, to extend it with a notion of
variable. Then, we have equipped this structure with the corresponding
monadic operation, \bind and \return. Whereas the \return is a trivial
variable former, the \bind operation corresponds to the substitution
of variables by terms. We have shown how to implement it, for any free
monad.

>>>>

The entire work of the magician stands here, in this extension. One
could be worried by the availability of such operations in our type
theory. Indeed, it would be inconvenient for the developer to have to
decide between two variants. The solution to this problem is, once
again, to use the type propagation system. Just as the $\F{switch}$
operator, $\F{switchD}$ can be automatically elaborated. To this end,
we extend the type-checking rules (Fig.~\ref{fig:type-checking}) with
the following inference rule:

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}$
completes the implementation. To sum up, we have implemented the free
monad construction for an arbitrary tagged description. This gives the
developer the ability, for any data-type, to extend it with a notion
of variable. Then, we have equipped this structure with the
corresponding monadic operation, \bind and \return. Whereas the
\return is a trivial variable former, the \bind operation corresponds
to the substitution of variables by terms. We have shown how to
implement it, for any free monad.

### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: 'user' evokes an interactive setting. Go back to the less frightening 'developer' Joe.

However, one could, quite rightfully, be worried by the introduction
of equalities in the picture. In a programming context, equalities
haunt us, begging for proofs. Hence, whereas the user typed a
constructor-constrained data-type, the elimination form will involve
equalities and their proofs. As much as possible, one would rather
avoid suffering from this burden. However, the equalities introduced
here are rather special: they are \emph{equations} constraining
variables to certain terms. Goguen et
al.~\cite{goguen:pattern-matching} have shown, in the context of
dependent pattern matching, how to tame these equations.

@@

Equations have simply disappeared. Note that it is straightforward to
turn this definition into a tagged form. We should precise that
forcing a description is not guaranteed to remove all constraints. It
is subject to future work to see if constraints can be entirely
eradicated, or presented more conveniently to the user. Finally, it is
worth mentioning that the forcing optimisation is a
\emph{source-to-source} transformation of the description.

>>>>

However, one could, quite rightfully, be worried by the introduction
of equalities in the picture. In a programming context, equalities
haunt us, begging for proofs. Hence, whereas the developer typed a
constructor-constrained data-type, the elimination form will involve
equalities and their proofs. As much as possible, one would rather
avoid suffering from this burden. However, the equalities introduced
here are rather special: they are \emph{equations} constraining
variables to certain terms. Goguen et
al.~\cite{goguen:pattern-matching} have shown, in the context of
dependent pattern matching, how to tame these equations.

@@

Equations have simply disappeared. Note that it is straightforward to
turn this definition into a tagged form. We should precise that
forcing a description is not guaranteed to remove all constraints. It
is subject to future work to see if constraints can be entirely
eradicated, or presented more conveniently to the developer. Finally,
it is worth mentioning that the forcing optimisation is a
\emph{source-to-source} transformation of the description.

### 2010-03-29 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: 'user' evokes an interactive setting. Go back to the less frightening 'developer' Joe.

However, our situation is significantly different from the one we
faced with encoded data: while we were suffering from a crucial lack
of information, we are now facing too much information, hence loosing
focus. This is a general issue with the usage of codes, as they convey
much more information than what the user is willing to see.

@@

Finally, and more importantly, we are able to present our user an
abstraction of \emph{finite} function space. This work as follow. When
an expression $\V{t}$ is pushed into a function which domain is a finite
set, we instead elaborate it as a term in a small $\pi$-type. That is,
a finite function. To realise our promise of a big $\Pi$-type, we
elaborate a call to the $\F{switch}$ eliminator.

The abstraction we have developed in this section are entirely
transparent to the user. In the case of finite function space, the
user has never to explicitly \emph{call} the $\F{switch}$
eliminator. For diagnosis purposes, a similar, reverse mechanism turns
low-level terms into high-level expressions. Hence, this system, which
complete presentation is out of the scope of this paper, seals the
low-level type theory behind the high-level expressions.

>>>>

However, our situation is significantly different from the one we
faced with encoded data: while we were suffering from a crucial lack
of information, we are now facing too much information, hence loosing
focus. This is a general issue with the usage of codes, as they convey
much more information than what the developer is willing to see.

@@

Finally, and more importantly, we are able to present the developer an
abstraction of \emph{finite} function space. This work as follow. When
an expression $\V{t}$ is pushed into a function which domain is a finite
set, we instead elaborate it as a term in a small $\pi$-type. That is,
a finite function. To realise our promise of a big $\Pi$-type, we
elaborate a call to the $\F{switch}$ eliminator.

The abstraction we have developed in this section are entirely
transparent to the user. In the case of finite function space, the
developer has never to explicitly \emph{call} the $\F{switch}$
eliminator. For diagnosis purposes, a similar, reverse mechanism turns
low-level terms into high-level expressions. Hence, this system, which
complete presentation is out of the scope of this paper, seals the
low-level type theory behind the high-level expressions.
